Binning Strategies and Related Loss for Large Data

### 2016-01-07 versionForJCGS/PaperJCGS.Rnw
Blind For Submission
Spell Check

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}
\newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
\newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Binning Strategies and Related Loss for Large Data}
\end{center}
  \medskip
} \fi

@@

The top row of Figure~\ref{fig:scatter-alpha} demonstrates several varients of the traditional scatterplot where each observation is plotted using distinct points. The traditional, solid point, scatterplot in Figure~\ref{fig:scatter-alpha}(a) shows a triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable. \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}(b)) to mitigate the problem of over-plotting. A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}(c)) which renders points as semi-transparent to provides more visibility of underlying points.
%All of these methods fall short in the example. As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help to a degree, but are also prone to over-plotting  when the data set is very large. Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. 
The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

@@

Other scatterplot adaptations have been introduced that avoid over-plotting by manipulating the display of the points by distorting the locations or the scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations, including those sharing identical coordinates, and use distortion of the point locations by having points repel one another to avoid overlapping. An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and resize the rows of cells according to density of points. This variable binning fragments the continuity of the axes into segments on different scales and also does not deal with points at identical coordinates. Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.

@@

Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot. A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size or 3D renderings to provide joint density information within each grid cell, known as a tile. A {\it sunflower plot} uses symbols that increase in complexity proportional to the number of points in each bin. Due to difficulty in rendering descernable shapes in limited space, sunflower plots are only useful when the number of points in each bin remains reasonably small. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.

@@

Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins. It is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies due to the improved handling of skewed bin count distibutions while maintaining similar frequency loss to standard frequency binning when four to seven frequency bins are used, as seen in Figure~\ref{fig:FreqLossCompare}. Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability. %In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.

@@

%----------------------------------------------------------------------------
\newpage
\begin{appendix}
\section{Appendix for Optimal Offset for Uniform Data Recorded to Resolution $\alpha$}  
%----------------------------------------------------------------------------
%\subsection{}
\label{proof:offset}

@@

$ b(x_i) = Q_x(.5) $ \\
$ \Rightarrow (x_1 - \theta) + (k\alpha/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
$ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha  $ \\
$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha  $ \\
$ \Rightarrow -2\theta = -\alpha $ \\
$ \Rightarrow \theta = \alpha/2 $ \\

>>>>

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}
\newcommand{\ktm}[1]{{\color{red} #1}} %red comments: Karsten
\newcommand{\hh}[1]{{\color{blue} #1}} %blue comments: Heike

@@

\if1\blind
{
\title{Binning Strategies and Related Loss for Large Data}
\author{Author A$^1$, Author B$^{2,3}$, Author C$^2$\\$^1$Affiliation X \\$^2$Affiliation Y \\$^3$Affiliation Z}
  \maketitle
} \fi

@@

The top row of Figure~\ref{fig:scatter-alpha} demonstrates several variants of the traditional scatterplot where each observation is plotted using distinct points. The traditional, solid point, scatterplot in Figure~\ref{fig:scatter-alpha}(a) shows a triangular structure is apparent with some outliers at a medium number of games and high number of strikeouts; however the density within the triangular mass of points is indistinguishable. \citet{tukey} suggested the use of open circles (see Figure~\ref{fig:scatter-alpha}(b)) to mitigate the problem of over-plotting. A modern alternative to open circles is alpha blending (see Figure~\ref{fig:scatter-alpha}(c)) which renders points as semi-transparent to provides more visibility of underlying points.
%All of these methods fall short in the example. As can be seen in Figure~\ref{fig:scatter-alpha}, strategy (a) is the least effective, as it provides information about the outliers and range of the data but cannot provide any point density information. Tukey's open circles (b) help to a degree, but are also prone to over-plotting  when the data set is very large. Alpha blending (c) highlights the structure, but minimizes the visual impact of outliers. 
The data set is large enough that neither alpha blending nor open circles are completely effective, and so we must pursue a different strategy which can provide better information about the relative density of points at a given location.

@@

Other scatterplot adaptations have been introduced that avoid over-plotting by manipulating the display of the points by distorting the locations or the scales. Generalized scatterplots \citep{Keim2010GenScatter} display all individual observations, including those sharing identical coordinates, and use distortion of the point locations by having points repel one another to avoid overlapping. An extension of generalized scatterplots uses clustering and local principal components to allow ellipsoid oriented distortion to display local correlation structure in the data \citep{Janetzko2013Ellipse}. Variable-binned scatterplots \citep{Hao2010visual} break the display into a non-uniform rectangular grid and re-size the rows of cells according to density of points. This variable binning fragments the continuity of the axes into segments on different scales and also does not deal with points at identical coordinates. Generalized and variable-binned scatterplots make fine data structure more visible and allow color to be reserved for a third variable instead of frequency; however, the distortion of the point locations and/or axes warp the visual display of the association between the two primary variables.

@@

Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot. A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size or 3D renderings to provide joint density information within each grid cell, known as a tile. A {\it sunflower plot} uses symbols that increase in complexity proportional to the number of points in each bin. Due to difficulty in rendering discernible shapes in limited space, sunflower plots are only useful when the number of points in each bin remains reasonably small. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above.

@@

Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins. It is strongly recommended to use the quantile-based algorithm to bin untransformed frequencies due to the improved handling of skewed bin count distributions while maintaining similar frequency loss to standard frequency binning when four to seven frequency bins are used, as seen in Figure~\ref{fig:FreqLossCompare}. Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability. %In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.

@@

%----------------------------------------------------------------------------
\newpage
\begin{appendix}
\section{Appendix for Origin Offset Proof}  
%----------------------------------------------------------------------------
%\subsection{}
\label{proof:offset}

@@

$ b(x_i) = Q_x(.5) $ \\
$ \Rightarrow (x_1 - \theta) + (k\alpha/2) = (x_{\lceil\frac{k+1}{2}\rceil}+x_{\lfloor \frac{k+1}{2}\rfloor})/2 $ \\
$ \Rightarrow 2x_1 - 2\theta + k\alpha = (x_1 + (\lceil\frac{k+1}{2}\rceil - 1)\alpha ) + (x_1 + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha ) $ \\
$ \Rightarrow -2\theta + k\alpha = (\lceil\frac{k+1}{2}\rceil - 1)\alpha  + (\lfloor\frac{k+1}{2}\rfloor - 1)\alpha  $ \\
$ \Rightarrow -2\theta + k\alpha = ((k+1) - 2)\alpha  $ \\
$ \Rightarrow -2\theta = -\alpha $ \\
$ \Rightarrow \theta = \alpha/2 $
